---
title: "Peer Assessment 1 - Reproducible Research"
date: "Saturday, March 14, 2015"
---

The following step will be to read the dataset into R Studio and load the necessary packages for this analysis
```{r read dataset}
dataset <- read.csv("ReproducibleResearch/activity.csv")
library(ggplot2)
library(dplyr)
```

Question 1: What is mean total number of steps taken per day?

Sub question 1.1: Calculate the total number of steps taken per day
```{r subquestion 1}
stepsbydate <- aggregate(steps ~ date, data = dataset, FUN=sum, na.rm=TRUE)
```

Sub question 1.2: Make a histogram of the total number of steps taken each day
```{r histogram}
hist(stepsbydate$steps, main="Total Steps by Date", xlab ="Total Steps")
```

Sub question 1.3: Calculate and report the mean and median of the total number of steps taken per day. 
Let's start with the mean:
```{r subquestion 3 - mean}
avgsteps <- aggregate(steps ~ date, data = dataset, FUN=mean, na.rm=TRUE)
print(avgsteps)
```

And the same for the median values:
```{r subquestion 3 - median}
mediansteps <- aggregate(steps ~ date, data = dataset, FUN=median, na.rm=TRUE)
print(mediansteps)
```

Question 2 is around the average daily activity pattern based on the time interval. The question is: What is the average daily activity pattern?

Subquestion 2.1:
Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken:
```{r time series plot}
avgdailypattern <- aggregate(steps ~ interval, data = dataset, FUN=mean)
timeseries <- ggplot(avgdailypattern, aes(x=interval, y=steps)) + geom_line()
print(timeseries)
```

The follow up question is, which interval contains the maximum number of steps:
```{r maxnum of steps}
maxinterval <- avgdailypattern[which.max(avgdailypattern$steps),]
print(maxinterval)
```

As there are many NA values in the dataset, we are looking at how to attach values to these, first it is important to find out how many NA values there are in the current dataset:
```{r numberofNAs}
sum(is.na(dataset))
```

The next step is to create a new dataset where the NA values are replaced by the mean value for the same interva:
```{r fill NAs}
newdataset <- dataset

avgsteps_perinterval <- newdataset %>% 
                   group_by(interval) %>%
                   summarise_each(funs(mean(steps, na.rm = TRUE)))
                 
avgsteps2 <- avgsteps_perinterval[,1:2]                 
                 
index_of_na <- which(is.na(newdataset$steps)) 

for (i in index_of_na) {
  newdataset$steps[i] <- with(avgsteps2, 
                                steps[interval == newdataset$interval[i]])
  
}
```

The following histogram presents the distribution of the total number of steps
```{r new histogram}
newtotaldailypattern <- aggregate(steps ~ date, data = newdataset, FUN=sum)
hist(newtotaldailypattern$steps, main = "Total Steps by Date" ,xlab =" Total Steps")
```

Then the mean and median total number of steps can be derived from the new dataset. Since the NA values are replaced by the mean, there is no difference between the first dataset and the new dataset with NAs replaced. Lets first look at the Mean values of the new dataset
```{r new mean}
newavgdailypattern <- aggregate(steps ~ date, data = newdataset, FUN=mean)
print(newavgdailypattern)
```

Then the new median values:
```{r new median}
newmediandailypattern <- aggregate(steps ~ date, data = newdataset, FUN=median)
print(newmediandailypattern)
```

To see the actual differences between the new dataset and the old dataset, the following table will present this:
```{r difference mean}
mergedavg <- merge(newavgdailypattern, avgsteps, by ="date", all=TRUE)
avgdiff <- mutate(mergedavg, StepsDifference = steps.x - steps.y)
print(avgdiff)
```

The same is done for the median values:
```{r difference median}
mergedmedian <- merge(newmediandailypattern, mediansteps, by ="date", all=TRUE)
mediandiff <- mutate(mergedmedian, StepsDifference = steps.x - steps.y)
print(mediandiff)
```

Finally, the impact of imputting missing data on the estimates of the total number of steps is as follows:
```{r difference sum}
mergedsum <- merge(newtotaldailypattern, stepsbydate, by="date", all=TRUE)
totaldiff <- mutate(mergedsum, StepsDifference = steps.x - steps.y)
print(totaldiff)
```


The last part of this assignment deals with the number of steps by day of the week. The first step for this analysis is to create new variable which indicates whether the date is a weekday or weekend:
```{r weekday or weekend}
newdataset$date <- as.POSIXct(newdataset$date)
newdataset$Weekday <- weekdays(newdataset$date, abbreviate = FALSE)
newdataset$DoWFlag <- ifelse(newdataset$Weekday == "Saturday" | newdataset$Weekday == "Saturday" , "Weekend", "Weekday")
newdataset$DoWFlag <- as.factor(newdataset$DoWFlag)
```

To see any difference between the two factors (Weekday or Weekend) a plot is created:
```{r day of week plot}
avgbydayofweek <- aggregate(steps ~ interval * DoWFlag, data = newdataset, FUN=mean)
lastplot <- ggplot(avgbydayofweek, aes(x=interval, y=steps)) + geom_line()
lastplot + facet_grid(DoWFlag ~.)
```

## The end!
